Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

The procedure $\conv$ in Figure~\ref{fig:encoding} translates the $\Sigma$-formula
$\varphi$ into an $\extendsig{\Sigma}$-formula.
This calls the subprocedure $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or
$\pnone$). It returns a pair $( t'\negvthinspace,\, D )$, where $t'$ is a term of
type $\tau$ and $D$ is a set of formulas encoding additional constraints.
%The role of $D$ is to ensure that
%the (restricted) function definition for $\con{f}$ includes certain tuples in
%its domain, which we explain more in the following.

@@

In the case that $t$ is an application of (interpreted) predicate $\boolop$ (e.g., a boolean connective, or equality),
we recursively call $\conv$ on its subchildren $t_i$ and polarity $\pol( \boolop, i, p )$, where $\pol$ is defined as\footnote{The negation $\neg p$ of polarity $p$ is 
$\pfalse$ if $p$ is $\ptrue$,
$\ptrue$ if $p$ is $\pfalse$, and
$\pnone$ if $p$ is $\pnone$.}:
\begin{equation*}
\pol( \boolop, i, p ) = \begin{cases}
                         \pnone & \boolop \equiv \teq \text{ or } ( \boolop \equiv \lite \text{ and } i=0 ) \\
                         \neg p & \boolop \equiv \neg \\
                         p & \text{otherwise}
                         \end{cases}
\end{equation*}
If $t$ has positive polarity,
we construct a conjunction of the result of the recursive calls with formulas $D_1, \ldots, D_n$.
Similarly, if $t$ has negative polarity,
we construct a disjunction of the result of the recursive calls with the negation of formulas $D_1, \ldots, D_n$.
It $t$ has no polarity,
then we return the union of $D_1 \cup \ldots \cup D_n$.

In the case that $t$ is a quantified formula $\forallf{\con{f}} \vec x.\; t_1$, it recursively calls $\conv$ on its body with the same polarity.
In the case that $t$ is a function definition of a function in $\sfundefs{\Sigma}$,
we instead construct a quantified formula over a single variable $y$ of type $\fargsort{\con{f}}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{\con{f}}( y )$.
In this case, since function definitions are a top-level conjunct, by case analysis on the return values of $\conv$, we know that $D_1$ is empty.
In the case that $t$ is not a function definition, 
the resulting quantified formula is reconstructed with the term $r_1$ returned by the recursive call,
and a quantified prefix is appended to the formula $D_1$ returned by the recursive call.

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature of linear arithmetic and uninterpreted functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$,
where $\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and 
$\sfuns{\Sigma_{\euf}} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}.\; \lite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is in definitional form with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

>>>>

The procedure $\conv$ in Figure~\ref{fig:encoding} translates the $\Sigma$-formula
$\varphi$ into an $\extendsig{\Sigma}$-formula.
This calls the subprocedure $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \emph{polarity} $p$, which is either $\ptrue$, $\pfalse$, or
$\pnone$. It returns a pair $( t'\negvthinspace,\, D )$, where $t'$ is a term of
type $\tau$ and $D$ is a set of formulas encoding additional constraints.
%The role of $D$ is to ensure that
%the (restricted) function definition for $\con{f}$ includes certain tuples in
%its domain, which we explain more in the following.

@@

If $t$ is an application of an interpreted predicate $\boolop
\in \{{\neg}, {\land}, {\lor}, {\teq}, {\lite}\}$,
we recursively call $\conv$ on the arguments $t_i$ and polarity $\pol( \boolop,
i, p )$, where $\pol$ is defined as
\begin{equation*}
\pol( \boolop, i, p ) =
\begin{cases}
\pnone & \text{if $\boolop = {\teq}$ or $(\boolop, i) = (\lite, 0)$} \\
\neg p & \text{if $\boolop = {\neg}$} \\
p & \text{otherwise}
\end{cases}
\end{equation*}
The negation $\neg p$ of polarity $p$ is $\pfalse$ if $p$ is $\ptrue$, $\ptrue$
if $p$ is $\pfalse$, and $\pnone$ if $p$ is $\pnone$.

If $t$ has positive polarity, we construct a conjunction of the result of the
recursive calls with formulas $D_1, \ldots, D_n$. Similarly, if $t$ has
negative polarity, we construct a disjunction of the result of the recursive
calls with the negation of formulas $D_1, \ldots, D_n$. It $t$ has no polarity,
then we return the union of $D_1 \cup \ldots \cup D_n$.

If $t$ is a quantified formula $\forallf{\con{f}} \vec x.\; t_1$, it
recursively calls $\conv$ on its body with the same polarity. If $t$ is a
function definition of a function in $\sfundefs{\Sigma}$, we instead construct
a quantified formula over a single variable $y$ of type $\fargsort{\con{f}}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{\con{f}}( y )$.
In this case, since function definitions are a top-level conjunct, by case
analysis on the return values of $\conv$, we know that $D_1$ is empty. In the
case that $t$ is not a function definition, the resulting quantified formula is
reconstructed with the term $r_1$ returned by the recursive call, and a
quantified prefix is appended to the formula $D_1$ returned by the recursive
call.

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature of linear arithmetic and uninterpreted functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$,
where $\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and 
$\sfuns{\Sigma_{\euf}} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}.\; \lite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation} 
%
Here, $s$ is a function that returns the sum of positive numbers between $0$
and its argument $x$. The formula $\varphi$ is in definitional form with
respect to functions $\sfundefs{\Sigma_u}$, and states that the sum of all
positive numbers between $0$ and $c$ is greater than $100$. This formula is
satisfiable with a model that interprets $c$ as an integer greater than or
equal to $14$. Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{s}$, and the uninterpreted function $\farg{s}$ of type $\fargsort{s}
\rightarrow \sortint$. The result of $\conv( \varphi, \ptrue )$ is the pair $(
\psi, \emptyset )$, where after simplification, $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{s}$, thus enabling finite model finding
techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{s}$. Moreover, all occurrences of $y$ in this formula are beneath
applications of the uninterpreted function $\farg{s}$, implying that the
formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09},
for which there exists a complete instantiation procedure. We observe that both
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second. $\square$
\end{example}
